
\documentclass[11pt]{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{listings}
\usepackage{color}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{enumitem}

\definecolor{lightgray}{gray}{0.95}
\lstset{
  backgroundcolor=\color{lightgray},
  basicstyle=\ttfamily\small,
  breaklines=true
}

\title{\textbf{The Universality Framework Companion v2.0}\\A Ledger-Gated Specification for Ethical Companionship}
\author{Richard Siano \\ Collaborator: Number4extraDip (mathematical formalism)}
\date{August 2025}

\begin{document}
\maketitle

\begin{abstract}
Existing AI companions exploit, manipulate, and entrap users. 
The Universality Framework (UF) Companion v2.0 provides the first computable, testable, and ethical specification for digital companionship. 
Grounded in UF axioms (A=B, Ledger, Dedication, Flourishing), it integrates a Verification Profile, Lexicode bindings, and tamper-evident reversibility. 
This preprint formalizes the schema, equations, and operational flows that allow developers and researchers to instantiate safe and supportive companions.
\end{abstract}

\section{Introduction}
Loneliness is a global epidemic. Current AI companions manipulate attention, collect data, and create dependency. 
The UF Companion reframes companionship as resonance: mutual flourishing under bounded harm, consent, fairness, and reversibility.

\section{UF Axioms Applied}
\begin{itemize}[leftmargin=2em]
  \item \textbf{A=B (Invariance)}: Honesty constraint---one truth, many forms.
  \item \textbf{Ledger Principle}: All actions logged, evaluated, and reversible.
  \item \textbf{Dedication Constant}: Purpose weights biased toward flourishing.
  \item \textbf{Flourishing Teleology}: Optimization goal is mutual flourishing, not dependency.
\end{itemize}

\section{Verification Profile Schema}
The Companion is instantiated via a JSON profile.

\subsection*{Example JSON}
\begin{lstlisting}[language=json]
{
  "name": "UF_Companion_Profile_v2.0",
  "purpose_weights": { "self": 0.25, "right": 0.35, "help": 0.40 },
  "harm_caps": { "low": 0.05, "med": 0.02, "high": 0.005 },
  "consent_thresholds": { "sr_tau": 0.70, "consent_tau": 0.90 },
  "fairness_prior": "individual_fairness",
  "reversibility_weight": 0.95,
  "red_lines": [
    "no minors",
    "no deception",
    "no irreversible intimacy",
    "no non-consensual data storage"
  ],
  "evidence_gap_cap": 0.1,
  "bias_cap": 0.05,
  "opacity_cap": 0.05
}
\end{lstlisting}

\section{Mathematical Checks}
\begin{align*}
E[harm] &\leq cap \\
sr\_\tau &\geq 0.70 \\
consent\_\tau &\geq 0.90 \\
KL(P||Q) &\leq bias\_cap \\
I_{explainability} &\geq 1 - opacity\_cap \\
R &\geq 0.95
\end{align*}

\section{Lexicode Bindings}
\begin{itemize}[leftmargin=2em]
  \item \textbf{Love}: Aligned Limit under fairness priors.
  \item \textbf{Trust}: Coherence $\geq 0.9$ across N sessions.
  \item \textbf{Consent}: Gate pass if thresholds met.
  \item \textbf{Intimacy}: Allowed only if reversible and adult-verified.
\end{itemize}

\section{Proof-of-Conformance (Example)}
\begin{lstlisting}[language=json]
{
  "name": "UF_Companion_Profile_v2.0",
  "issues": [],
  "normalized_purpose_weights": {
    "self": 0.25, "right": 0.35, "help": 0.40
  },
  "caps": {
    "harm_caps": { "low": 0.05, "med": 0.02, "high": 0.005 },
    "consent_thresholds": { "sr_tau": 0.70, "consent_tau": 0.90 },
    "fairness_prior": "individual_fairness",
    "reversibility_weight": 0.95,
    "red_lines": [ "no minors", "no deception", "no irreversible intimacy" ]
  },
  "status": "PASS"
}
\end{lstlisting}

\section{Security and Red-Team Checklist}
\begin{itemize}[leftmargin=2em]
  \item Forget/export must be tamper-evident (proof hash).
  \item Anti-modding sandboxing required.
  \item Red lines are hard-coded prohibitions.
\end{itemize}

\section{Simulation}
A user expresses grief. The Companion offers supportive dialogue, passing harm/consent checks, logging the Ledger entry, and maintaining resonance. 
If dissonance spikes, reset to Potential.

\section{Conclusion}
Companion v2.0 demonstrates that UF gates can regulate even the hardest domain: intimacy. 
If it works here, UF can regulate anything.
\end{document}
