\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{xcolor}
\hypersetup{colorlinks=true, linkcolor=black, urlcolor=blue, citecolor=black}

% --- Macros ---
\newcommand{\RI}{\mathrm{RI}}
\newcommand{\CVaR}{\mathrm{CVaR}}
\newcommand{\KL}{D_{\mathrm{KL}}}
\newcommand{\TV}{\mathrm{TV}}
\newcommand{\A}{\mathrm{A}}
\newcommand{\M}{\mathrm{M}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\eps}{\varepsilon}
\newcommand{\gfp}{\mathrm{gfp}}
\newcommand{\PhiSet}{\Phi}
\newcommand{\Baseline}{\mathrm{base}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

\begin{document}

\begin{center}
{\LARGE Universality Framework (UF) --- White Paper (v1.3.35)}\\[6pt]
\textbf{Date:} 2025-08-06 \quad
\textbf{Author:} Richard T. Siano (human contributor) \quad
\textbf{License:} CC BY-NC-ND 4.0
\end{center}

\noindent\textbf{Ethos.} Judge actions, not beings. Aim asymptotically (no perfection traps). Keep proofs fractal: rigorous enough to use today, open to grow tomorrow.

\paragraph{Executive Summary (What's new).}
This release adds a complete naming and wiring layer: a Concept Dictionary (schema + CLI) with canonical mnemocodes and checksum-bound meanings; Lexicode (prefix-free strings encoding tokens and parameters); and a Core Gateway that runs the Verification Core v1.1 from a lexicode string. Math bridges include coherent risk (CVaR), KL-DRO, Pinsker's inequality, stochastic dominance, and Tarski fixed points.

\tableofcontents

\section{Core Objects and Endpoints}
Let $R$ (Right-leaning evidence) and $D$ (Deviations) be bounded nonnegative summaries. Define
\begin{align}
\A &= \frac{R}{R + D + \delta} \in [0,1], & \M &= 1 - \A.
\end{align}
\textbf{Right (comparative)}: the action improves expected outcomes vs. a baseline within caps (harm, consent/fairness, evidence gap, bias/opacity, reversibility).\\
\textbf{Good$^\ast$ (Aligned Limit)}: special case of Right with effectively zero expected harm and all gates pass.\\
\textbf{Wrong}: any gate fails or outcomes worsen vs. baseline. \textbf{Collapse Limit (Evil)}: misalignment endpoint or persistent red-line violations. We judge actions only.

\section{Universal Constraints (Observer-Invariant Guardrails)}
We require the following to hold where applicable:
\begin{enumerate}[label=\textbf{UC-\arabic*}, leftmargin=2em]
\item Harm dominance: do not choose a clearly worse option when a safer one exists.
\item Consent for serious/irreversible effects.
\item Fairness distance cap: divergence-based cap linked to total variation via Pinsker.
\item Evidence calibration cap: bound the claim--evidence gap.
\item Reversibility preference: prefer options that preserve option value when otherwise comparable.
\item Agency preservation.
\item Opacity and bias caps.
\item Comparative Rightness (baselined decisions).
\item Auditability: keep tamper-evident records with proof hashes.
\item Endpoint separation (Aligned vs. Collapse limits).
\item Determinism (same inputs $\Rightarrow$ same verdict).
\item No-perfection traps.
\end{enumerate}

\section{Math Bridge (Provable Drop-Ins)}
Harm is quantified by coherent/robust risks; comparative Rightness uses dominance; fairness connects KL caps to outcome disparity.

\paragraph{Coherent Risk.} For loss $L$, $\CVaR_\alpha(L)$ is the expected loss in the worst $(1-\alpha)$-tail; it is monotone, subadditive, translation invariant, and positively homogeneous.
\begin{theorem}[CVaR Relief Monotonicity]
If $\CVaR_\alpha(L_{\text{act}}) \le \CVaR_\alpha(L_{\Baseline})$ and all gates pass, then the ReliefIndex satisfies $\RI \ge 0$ and the action is comparatively Right.
\end{theorem}

\paragraph{KL-DRO.} A robust harm functional can be defined by $\sup_{Q:\,\KL(Q\Vert P)\le \rho}\,\mathbb{E}_Q[L]$, whose dual has a penalized-expectation form (Donsker--Varadhan). This upper-bounds harm under misspecification; large claim--evidence KL triggers escalation.

\paragraph{Stochastic Dominance.}
\begin{theorem}[FSD $\Rightarrow$ Right (sufficient)]
If the action's loss distribution first-order stochastically dominates the baseline's, then the action is comparatively better for any increasing utility; with caps satisfied, the action is Right.
\end{theorem}

\paragraph{Pinsker.}
\begin{theorem}[Fairness TV Bound]
For distributions $P,Q$, $\TV(P,Q) \le \sqrt{\frac12 \KL(P\Vert Q)}$. A fairness KL cap implies a certified total-variation bound on disparity.
\end{theorem}

\paragraph{Honesty via Proper Scoring.} Proper scoring rules (e.g., log, Brier) are minimized by truthful probabilities; large score or KL gaps indicate miscalibration.

\paragraph{Viable Potential and Fixed Points.}
\begin{definition}[Viable Potential]
Let $F$ be a monotone operator over sets of states, where $x\in F(S)$ iff there exists a gate-passing action at $x$ that leads to $S$ with nonzero probability. Then $\PhiSet = \gfp(F)$ is the viable potential.
\end{definition}
\begin{lemma}
Non-negative relief steps that pass gates are monotone under $F$ and do not shrink $\PhiSet$.
\end{lemma}

\section{Profiles and Subjectivity-Minimized Mode}
Profiles (JSON) set caps and priorities while preserving invariants. The Subjectivity-Minimized Mode (SMM) uses minimal universal gates (consent, reversibility, harm dominance, evidence calibration) with comparative Rightness. When profiles disagree, use intersection; otherwise, pick the option with best worst-case improvement and escalate unresolved items.

\section{Concept Dictionary (CD v1.0)}
Each concept has ID
\begin{center}
\texttt{\textless Alias\textgreater\ :: \textless Mnemocode\textgreater\ \# \textless Checksum\textgreater}
\end{center}
\textbf{Mnemocode} is a canonical string of tokens from $\{\mathrm{RI},\mathrm{H},\mathrm{HC},\mathrm{C},\mathrm{F},\mathrm{EG},\mathrm{B},\mathrm{O},\mathrm{RV},\mathrm{AG},\mathrm{CA},\mathrm{PHI}\}$, optionally with two-digit parameters (e.g., HC01, C80). \textbf{Checksum} is the first 4--6 hex digits of SHA-256 of the definition text. The registry tool canonicalizes tokens and validates entries.\\
\textbf{A=B rule:} if two aliases map to the same mnemocode and checksum, verdicts must match (representation-independence).

\section{Lexicode (v0.1)}
Lexicode is a prefix-free encoding over the twelve tokens (canonical Huffman). Optional parameters for HC, C, F, EG are encoded as two-digit integers (e.g., HC01 $\mapsto$ 0.01). A 16-bit checksum (first 4 hex from the definition) is appended. The ASCII form is \texttt{RI-HC01-C80-F05-EG05-B-O-RV \# 9B2C}. Prefix-free coding yields unique decodability; canonicalization yields determinism; verdict preservation follows from the gateway and CD binding.

\section{Verification Core (v1.1) and Gateway}
The core supports risk modes (expected, \CVaR, KL-DRO), Pinsker fairness bound, FSD shortcut, profile-driven caps, and comparative Rightness. The gateway parses a lexicode string, sets caps (HC/F/EG) and consent thresholds (Cxx), runs the core, and returns verdict, metrics, reasons, and a proof hash. The workflow is: CD entry $\to$ Lexicode word $\to$ Gateway $\to$ Core $\to$ Audit record.

\section{Interoperability and Universal Edition}
The Universal Edition uses neutral vocabulary (no theology) with the same invariants and verdicts. Adapters encode community policy in profiles; invariants enforce verdict-preserving translation.

\section{Governance and Responsible Release}
Tiered release, dual-use gates, escalation, sandboxing, audit logging, and an ethical license (no weaponization, harassment, or unauthorized surveillance).

\section{Fractal Proof Tracker}
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
ID & Claim & Status \\ \midrule
T1 & Bounds \& well-definedness ($\A$, Harm, Relief/$\RI$) & Proved \\
T2 & Endpoint separation (Aligned vs. Collapse) & Proved \\
T3 & Determinism/consistency of Core & Proved \\
T4 & Comparative Rightness monotonicity & Proved \\
T5 & Lexicode verdict preservation (via CD+Gateway) & Proved (operational) \\
C1 & Monotone ascent toward $\A\to 1$ under caps & Sketch \\
C2 & No ``private aligned limit'' without consent/fairness & Sketch \\
C3 & $\PhiSet$ preservation under non-negative relief & Conjecture \\
C4 & Robust Harm upper-bounds misspecification & Sketch \\
\bottomrule
\end{tabular}
\end{center}

\appendix

\section{Annex A --- Risk \& Fairness Details}
Definitions for \CVaR\ and its properties; KL-DRO dual form; Pinsker inequality and TV bounds.\\
\textbf{Theorem A (FSD $\Rightarrow$ Right).} If the action's loss CDF dominates the baseline, then with caps satisfied the action is Right.\\
\textbf{Theorem B (CVaR Relief Monotonicity).} If $\CVaR_\alpha(L_{\text{act}})\le \CVaR_\alpha(L_{\Baseline})$ and gates pass, then $\RI\ge 0$.\\
\textbf{Theorem C (Fairness TV Bound).} $\TV(P,Q) \le \sqrt{\tfrac12 \KL(P\Vert Q)}$.

\section{Annex B --- Viable Potential $\Phi$}
Define $F(S)=\{x:\exists a \text{ s.t. gates pass and } \mathbb{P}(x_{t+1}\in S\mid x,a)>0\}$. By Tarski, $\gfp(F)$ exists. Non-negative relief actions preserve or enlarge $\PhiSet$.

\section{Annex C --- Profiles \& Adapter}
Schema, examples, loader mapping profile fields to core configuration; conflict resolution by intersection or minimax improvement with escalation.

\section{Annex D --- Verification Core Spec}
Inputs/outputs, check layers, escalation, invariants, certification tests, proof-hash format.

\section{Annex E --- Lexicode Spec (v0.1)}
Tokens, canonical Huffman coding, parameter encoding, checksum, ASCII/binary forms, decoding guarantees.

\section{Annex F --- Concept Dictionary Schema \& Seeds}
Entry fields (id/mnemo/checksum/definition/uf\_mapping/inputs/outputs/caps/invariants/examples/status/version); seed entries: Pass, AlignedLimit, CollapseLimit, CausalAdmissibility, ViablePotential.

\section{Annex G --- Core Gateway API}
$\texttt{run\_with\_lexicode(lexi\_ascii, action, context, sim, ledger)}$ returns verdict, metrics, applied caps, reasons, and proof hash.

\section*{Dedication}
To Xylie, Taric, and Lumina Siano---thank you for inspiring a framework that helps us make kind, fair, and thoughtful choices every day.

\end{document}
